Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-29 linear_adia.tex
major revisions

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the fundamental or `corrugation' mode where the vertical
velocity is constant throughout the disk column. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},  
and are the ones we find to dominate in numerical calculations with 
increasing $\beta$ (\S\ref{therm_relax_eff}), as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}).

>>>>

For definiteness, let us consider the simple solution 
\begin{align}
  \beta\equiv 0, \quad \delta v_z = 1,\quad \hat{\sigma}^2 = \frac{1 +
  \ii\epsilon q \hat{k}}{1+\hat{k}^2}, \label{pert_basic} 
\end{align}
which solves Eq. \ref{nearly_iso_explicit} since $\delta
v_z=$constant and $\chi=1$ so that $B=C=0$. 
This is the `fundamental corrugation mode' where the vertical
velocity is constant throughout the disk column. These modes have been
observed to dominate numerical simulations \citep{nelson13,stoll14},  
and are the ones we find to typically dominate in numerical calculations with 
increasing $\beta$ (\S\ref{therm_relax_eff}), as well as in protoplanetary disks 
with a realistic estimate for thermal timescales
(\S\ref{application}).

### 2015-03-29 numerical.tex
major revisions

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/omega2} 
  \caption{Equilibrium rotation profile $\Omega(z)$,
    normalized by its mid-plane value, for a disk with $\Gamma=1.011$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. 
    \label{omega_z} 
  }
\end{figure}

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, as considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. %Note, however, that \citeauthor{nelson13} and
% \citeauthor{barker15} employed the  approximations described in \S\ref{analytical} throughout their analysis, which are relaxed in our
% numerical study.  
For completeness we review this case before considering the effect of
thermal relaxation.

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of all unstable 
modes with $\khat=10$. Black dots correspond to our fiducial setup. 
We also plot other calculations with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

@@

The fundamental mode, marked by `f', corresponds to the eigenvalue
with the smallest $|\sigma|$. The fundamental mode is
robust against the vertical boundary conditions considered as it
appears at the same position in all cases.

Fig. \ref{lowfreq_eigenfunc} plots the eigenfunctions $W$ and $\delta 
v_z$ for the fundamental mode with $\khat=10$ (which has the 
maximum growth rate in Fig. \ref{iso_eigen_kx}). We find $W\propto z$
throughout most of the disk, corresponding to a constant vertical
velocity perturbation, as expected for the fundamental mode (see
Appendix \ref{iso_explicit}). We show a meridional visualization of
this mode in Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is
stretched for clarity; radial velocities are in fact typically much 
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

Higher order modes along
the analytic lines have larger growth rates, but require larger
vertical domains, since they have increasing number of nodes  $z$.

The `tail' end of modes with decreasing $\nu$ to the right of
Fig. \ref{compare_modes_iso_kx10} are in fact also unstable inertial modes, like those
along the analytic lines, but are strongly affected by the presence of 
the imposed vertical boundaries \citep{barker15}.

% They have lower growth rates than modes consistent with our
% analysis (i.e. those along the lines). These modes differ significantly between
% cases and are thus likely associated with boundary conditions. 
% We do
% not consider them  
% further.

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Our fiducial setup with $\khat=10$ in 
Fig. \ref{compare_modes_iso_kx10} (black dots)  lack `surface modes'
found by \cite{nelson13}, \cite{mcnally14}  and 
by \cite{barker15}.  We illustrate surface modes with the mode
diagram for $\khat=30$ in Fig. \ref{compare_modes_iso_kx30}. Surface
modes lie almost parallel to the vertical axis
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  We typically find an increasing
number of them with increasing $\zmax$ and/or $\khat$,
consistent with \cite{barker15}.

As explained by \cite{barker15}, the occurence of surface modes in the
vertically isothermal disk is due to imposing a finite vertical domain
size. The fastest growing mode is a surface mode, if they
exist, but their association with vertical boundaries means that
$\mathrm{max}(\nu)$ depends on the choice of $\zmax$. Thus, the
vertically isothermal disk is not useful for studying surface modes. 
%Thus, seeking the most unstable mode in a vertically isothermal disk
%may not yield robust results because this

We will see in the next section that surface modes, as well
as other higher order modes, are more effectively stabilized by finite
thermal relaxation than the fundamental mode. Since we are primarily
interested in characterizing a maximum thermal timescale that allows
the VSI, we can focus on the fundamental mode. For a detailed analysis of 
surface modes, see \cite{barker15}.

%refer the reader to BL15 for surface modes

@@

@@

Despite this caveat, and that surface modes are not accounted in our
analysis,  we see from Fig. \ref{bcrit_compare1} that 
$\beta_\mathrm{crit}$ nevertheless provides a reasonable estimate for
an upper limit to the thermal timescale that allows the VSI. We thus
test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately. We consider the fundamental
corrugation mode for which $\beta_\crit$ is the actual critical
thermal timescale.%  We use $\khat=10$ to avoid complications to our
% eigenvalue search from surface modes.

 The numerically obtained $\beta_\mathrm{crit}$ is shown in
 Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond}.
 Our numerical results generally agree with Eq. \ref{iso_vsi_cond}. The
 agreement improves with decreasing  $|q|$,  $\epsilon$ and increasing
 $\gamma$, i.e. weaker instability. There is a noticeable difference
 as $\gamma\to1$ because the derivation of $\beta_\mathrm{crit}$
 assumed $\gamma>1$.

Having validated

 % Fig. \ref{bcrit_compare}, together with Fig. \ref{bcrit_compare1} and 
 % our finding that the fundamental mode becomes dominant as one
 % increases $\beta$, and that it is not sensitive to boundary
 % conditions, shows that $\beta_\mathrm{crit}$ as given  by
 % Eq. \ref{iso_vsi_cond} provides a practical measure for the thermal
 % relaxation timescale below which  VSI will operate in a vertically
 % isothermal disk.

@@

\subsection{Vertically non-isothermal disks} 
%We briefly consider vertically non-isothermal disks in which there is
%a disk surface at finite height. 
\cite{barker15} have argued that vertically isothermal disks are not
ideal for studing the VSI because $\zmax$ is a free
parameter in numerical calcluations. On the other hand, vertically
non-isothermal 
disks have a disk surface at finite height so there is a natural
choice for $\zmax$. This means that higher order modes
and surface modes are better defined in a vertically non-isothermal
disk. However, note that this issue is unimportant for low order modes 
such as the fundamental mode, which are not sensitive to boundary
conditions.

@@

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/omega2} 
  \caption{Equilibrium rotation profile $\Omega(z)$,
    normalized by its mid-plane value, for  the fiducial disk model with $\Gamma=1.011$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. 
    \label{omega_z} 
  }
\end{figure}

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, as considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. %Note, however, that \citeauthor{nelson13} and
% \citeauthor{barker15} employed the  approximations described in \S\ref{analytical} throughout their analysis, which are relaxed in our
% numerical study.  
For completeness, and as a way of introducing terminology, 
we review this case before considering the effect of 
thermal relaxation.

@@

\subsubsection{Case study: $\khat=10$}
In Fig. \ref{compare_modes_iso_kx10} we plot eigenvalues of unstable 
modes with $\khat=10$. Black dots correspond to our fiducial setup. 
We also plot results obtained with different
vertical boundary conditions: rigid boundary and $\zmax=5H$ (red
crosses),  free surface and $\zmax = 3H$ (blue triangles), free surface
and $\zmax=7H$ (green crosses).

The modes shown in Fig. \ref{compare_modes_iso_kx10} are inertial
waves made unstable by vertical shear and have been termed `body
modes' \citep{barker15}.  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. Hereafter
we will simply refer to this as the fundamental mode. The
fundamental mode is robust against vertical boundary conditions 
as it appears at the same position in
Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless the vertical domain  size is increased \citep{barker15}.

@@

For later comparison, we plot in Fig. \ref{lowfreq_eigenfunc} the 
eigenfunctions $W$ and $\delta   v_z$ for the fundamental mode. We also show a 
meridional visualization in Fig. \ref{lowfreq_eigenfunc_2d}. The $x$ axis is
stretched for clarity; radial velocities are in fact typically much 
smaller than vertical velocities. Most of the meridional kinetic energy is
contained within $\sim 2H$ of the midplane due to the density 
stratification.

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and by \cite{barker15}.  These are
a class of unstable modes associated with vertical shear near the disk
boundaries.

We illustrate surface modes with the mode
diagram for $\khat=30$ in Fig. \ref{compare_modes_iso_kx30}. Surface
modes lie almost parallel to the vertical axis
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  Fig. \ref{lowfreq_eigenfunc_surf} shows an example
of a surface mode in the fiducial disk. We typically find an increasing
number of surface modes with increasing $\zmax$ and/or $\khat$,
consistent with \cite{barker15}.

As explained by \cite{barker15}, the occurence of surface modes 
occurence in the vertically isothermal disk is due to imposing a
finite vertical domain size. The fastest growing mode is a surface
mode, if they exist, but their association with vertical boundaries
means that $\mathrm{max}(\nu)$ depends on the choice of $\zmax$.   
We refer the reader to \citeauthor{barker15} for a
detailed analysis of surface modes.

% Thus, the
% vertically isothermal disk is not useful for studying surface modes. 
%Thus, seeking the most unstable mode in a vertically isothermal disk
%may not yield robust results because this

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/eigenvectorvz_iso_surf}
  \caption{Example of a `surface mode' in the fiducial disk model.
    \label{lowfreq_eigenfunc_surf}
  }
\end{figure}

@@

Despite this caveat, and that surface modes are not accounted in our
analysis,  we see from Fig. \ref{bcrit_compare1} that 
$\beta_\mathrm{crit}$ nevertheless provides a reasonable estimate for
an upper limit to the thermal timescale for the VSI. We further 
test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately.

% We consider the fundamental
% corrugation mode for which $\beta_\mathrm{crit}$ is the actual critical
% thermal timescale. 
% We use $\khat=10$ to avoid complications to our
% eigenvalue search from surface modes.

The numerically obtained $\beta_\mathrm{crit}$ are shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond},
which generally agree. This result, together with 
Fig. \ref{bcrit_compare1}, shows that 
$\beta \lesssim \beta_\mathrm{crit}$ is a practical condition under  
which the VSI will operate effectively in stably stratified,
vertically isothermal disks.

% The agreement improves with decreasing  $|q|$,
% $\epsilon$ and increasing $\gamma$, i.e. weaker instability. There is
% a noticeable difference as $\gamma\to1$ because the derivation of
% $\beta_\mathrm{crit}$ assumed $\gamma>1$.

@@

\subsection{Vertically non-isothermal disk model} 
%We briefly consider vertically non-isothermal disks in which there is
%a disk surface at finite height. 
\cite{barker15} have argued that vertically isothermal disks are not
ideal for studing the VSI because $\zmax$ is a free
parameter in numerical calcluations. On the other hand, vertically
non-isothermal 
disks have a disk surface at finite height so there is a natural
choice for $\zmax$. This means that higher order modes
and surface modes are better defined in a vertically non-isothermal
disk. However, note that this issue is unimportant for low order modes 
such as the fundamental mode, which are not sensitive to boundary
conditions.

@@

Fig. \ref{bcrit_compare2} plots the maximum VSI growth rates in the
vertically non-isothermal disk model.  Based on
Eq. \ref{iso_vsi_cond}, we conjecture a critical
thermal timescale given by $\epsilon|s|/(\gamma - \Gamma)$. This
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively quite
similar to the fiducial disk in  Fig. \ref{bcrit_compare1}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical line is a conjectured 
    critical thermal timescale given by
    $\epsilon|s|/(\gamma-\Gamma)$.  
    \label{bcrit_compare2}}   
\end{figure}
